Computing likelihoods under the SMC for a general class of ARGs.

### 2024-04-25 paper.tex
Update notation.

First, we define the SMC process in a ``backwards-in-time'' way,
rather than the usual ``along-the-genome'' way.
Note that this is a
reiteration of the original work of \citet{mcvean_approximating_2005}.
Then, we describe
in more detail what sort of information we have in estimated ARGs.
In the subsequent section we connect the two and
show how the ``backwards-in-time'' description
makes it relatively straightforward to compute the likelihood
for a more general class of estimated ARGs.
Finally, we close with a demonstration that the likelihood computation can be used
to do Bayesian inference on ancestral population sizes.

@@

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.
Although \citet{mcvean_approximating_2005} verbally described the SMC 
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend 
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.

We now define the process more formally,
and following the notation of \citet{mcvean_approximating_2005}.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each ``lineage'' $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
genetic material inherited by either one of a set of samples.
This ancestral material \citep{wiuf_ancestry_1999} carried by a
lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
which each are given new, unique labels
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
The newly formed lineage acquires the union of both intervals, % L is updated accordingly
and the original lineages are removed.
As a consequence, under the SMC common ancestry events are always
associated with coalescence events
along some fraction of the merging half-open intervals.

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
definition of the strict total order it follows that
if $I_{jk} = 1$ then $I_{kj} = 0$).

For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations.
In particular,
although recombination affects \emph{which} lineages can coalesce into $j$,
it does not change the \emph{number} of such lineages:
in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.
Another useful quantity to know is the number of lineages carrying material
at each point on the genome:
for a location $z$ and time $t$, this is
\begin{equation}
    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
\end{equation}

After any coalescence or recombination event,
the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
are replaced by a single segment $[x, y)$ whose endpoints have not fully
coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$.
By reasoning about the process in this way we can keep representing 
all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
and the process terminates.

[TO DO: remark on  beyond local roots of the genealogy]

@@

Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals (that we call \emph{samples}).
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

\begin{figure}[!ht]
\centering
% JK: I put in 0.9 here because the fig is too big and takes up more than 
% a page. This flushes all the figs to the end of the document. We should 
% either make it a bit smaller or reduce the caption.
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

A key observation of \citet{mcvean_approximating_2005}
was that the sequence of trees along the genome
generated by the SMC was Markovian \emph{after removing all unary nodes}.
In other words, if we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes,
then the distribution of the next tree can be generated only knowing the current tree.
(To see that the unary nodes makes the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
However, removing these unary portions of nodes seems odd
when one considers the consequences of the inheritance paths through the
ARG in the top figure.
As pictured, both figures tell us that
3 inherits the middle bit of the genome from 7 through 5.
However, if we erase 5 from the second tree,
does this suggest that 3 inherited a single segment from 7 along two distinct paths?
The haplotype inheritance graph above makes it clear that this is unlikely:
it is in fact reasonable to leave in these unary sections.

Furthermore, these locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
[TODO ADD SINGER]
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

% % % % % % % % % % % % % % % % % %
\subsection{Calculating likelihoods} \label{par:liks}

@@

More specifically,
let $\deg_o(n) = \#\{e \in E : p_e=n\}$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ common ancestry events,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Note however that this description is general and holds for the case of
more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
contain polytomies. By treating these as if there were zero-length edges
between a sequence of mergers; the order of these unresolved mergers does 
not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
This is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.

Similarly,
let $\deg_i(n) = \#\{e \in E : c_e=n\}$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/single_event.png}
    \caption{
        Proposed figure to explain the likelihood computation with.
        TODO: change numbered labels to match the order in the text,
        add $x$ and $y$.
        \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times A)$, where $r$ is the recombination rate
and $A$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occurr, then the number of links not
affected by recombination is 1 less (see \ref{eq:depth}).

>>>>

First, 
we describe
in more detail what sort of information we have in estimated ARGs.
Then,
we define the SMC process in a ``backwards-in-time'' way,
rather than the usual ``along-the-genome'' way.
Note that this is a
reiteration of the original work of \citet{mcvean_approximating_2005}.

In the subsequent section we connect the two and
show how the ``backwards-in-time'' description
makes it relatively straightforward to compute the likelihood
for a more general class of estimated ARGs.
Finally, we close with a demonstration that the likelihood computation can be used
to do Bayesian inference on ancestral population sizes.

Although the term Ancestral Recombination Graph (ARG) has been used to mean several things,
in the generic sense, it is a labeled graph structure that records
the inheritance of genetic material \citep{wong_general_2023}.
An ARG of the sort we consider describes the history of inheritance
of the genomic material of a focal set of individuals, or \emph{samples}.
Concretely, an ARG is equivalent to a collection of (non-contradictory) statements
of the form ``$c$ inherited from $p$ on the genomic segment from $x$ to $y$'',
where $c$ and $p$ are (haploid) genomes, and $x$ and $y$ are genetic coordinates.
We record each such statement in an ``edge'',
often summarized as the tuple $(p,c,x,y)$.
By ``describes the history'',
we mean that each node in the ARG represents a specific (ancestral) genome,
and that the ARG specifies the genealogical tree describing how the samples are related
at each point on the genome.

\begin{figure}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_custom_2rows_area_full_hap.png}
\caption{ARG (top), generated under the SMC, and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondence
between both representations a node is recorded for each local tree whenever
a node is encountered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows show the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
By removing recombinant nodes, we limit the information on the time to each
of these events to the windows (blue shaded array) delimited by by the age of
the child node $c$ and the minimum age of the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

%% Noved this section to section on smc
% A key observation of \citet{mcvean_approximating_2005}
% was that the sequence of trees along the genome
% generated by the SMC was Markovian \emph{after removing all unary nodes}.
% In other words, if we remove recombinant nodes entirely,
% and erase the unary portions of coalescent nodes,
% then the distribution of the next tree can be generated only knowing the current tree.
% (To see that the unary nodes makes the process non-Markov,
% consider that a node in a tree might be unary either because it's about to be a coalescent node
% or it already has been:
% looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
% because it was not present in the first tree.
% So, the first tree gives us information about subsequent trees
% that the second tree alone could not.)
% However, removing these unary portions of nodes seems odd
% when one considers the consequences of the inheritance paths through the
% ARG in the top figure.
% As pictured, both figures tell us that
% 3 inherits the middle bit of the genome from 7 through 5.
% However, if we erase 5 from the second tree,
% does this suggest that 3 inherited a single segment from 7 along two distinct paths?
% The graph-view of the ARG (above) makes it clear that this is unlikely:
% it is in fact reasonable to leave in these unary sections.

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014},
\tsinfer{} \citep{kelleher_inferring_2019},
\kwarg{} \citep{ignatieva_kwarg_2021}, and
\texttt{SINGER} \citep{deng2024robust} estimate ARGs 
including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineages with \emph{overlapping} genomic intervals may coalesce.
By limiting common ancestry events to overlapping pairs of lineages, we are
guaranteed that all nodes in the ARG only carry ancestral material, this is
genetic material that is inherited by either one of a set of samples \citep{wiuf_ancestry_1999}.
As such, we only require a single half-open interval to describe the ancestral
material associated with each common ancestor node in the ARG.
Surprisingly, although the SMC can be defined
in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.
Linking this back to relationship between a graph and local trees view on
the ARG, then the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
(To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because it's about to be a coalescent node
or it already has been:
looking at the second tree (Fig.\ref{fig:smc-unary}), we know node 5 will be coalescent further down the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.)
Although \citet{mcvean_approximating_2005} verbally described the SMC
using this backwards-in-time description, their analysis
was almost entirely in left-to-right terms,
and subsequent papers followed this trend
\citep{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014,rasmussen_genome-wide_2014}.

We now define the process more formally,
and following the notation of \citet{mcvean_approximating_2005}.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Since we look backwards in time, $t=0$ is today, and so
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
which each are given new, unique labels and added to $L(t)$
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$. From the
definition of the strict total order it follows that
if $I_{jk} = 1$ then $I_{kj} = 0$.

For future use, define $C_j(t)$ to be the set of lineages that can coalesce into lineage $j$,
i.e., $C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$. And therefore, 
$|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$.
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations.
In particular,
although recombination affects \emph{which} lineages can coalesce into $j$,
it does not change the \emph{number} of such lineages:
in other words, a recombination event changes $C_j(t)$ but not $I_{j}(t)$.

%% I have updated the description of the process below to avoid having to
% define yet another symbol.
%Another useful quantity to know is the number of lineages carrying material
%at each point on the genome:
%for a location $z$ and time $t$, this is
%\begin{equation}
%    N(t,z) = \vert\{j : z \in X_j(t) \} \vert
%\end{equation}

To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

%After any coalescence or recombination event,
%the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
%are replaced by a single segment $[x, y)$ whose endpoints have not fully
%coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
%and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
%are not added back to $L(t)$.

Finally, by describing the SMC
backwards in time, some intervals contained within the half-open
interval representing a lineage might have fully coalesced before all the genetic
material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
intervals however do not affect the marginal genealogies.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Calculating likelihoods} \label{par:liks}

@@

More specifically,
let $\deg_o(n) = \vert\{e \in E : p_e=n\}\vert$ be the out-degree of a node $n$,
i.e., the number of edges having $n$ as a parent.
Since mergers are binary in the SMC,
this must have been the result of $\deg(n) - 1$ common ancestry events,
and so this contributes a factor of $\lambda^{\deg_o(n)-1}$ to the likelihood
for nodes with any children.
Note however that this description is general and holds for the case of
more-than-binary mergers. Indeed, inferred ARGs (e.g.\ \texttt{tsinfer}) may
contain polytomies. By treating these as if there were zero-length edges
between a sequence of mergers; the order of these unresolved mergers does 
not affect the likelihood. Under the SMC the likelihood of a zero-length edge is zero.
This is not special: since edge lengths are continuously distributed,
the likelihood of any particular value is also zero.

Similarly,
let $\deg_i(n) = \vert\{e \in E : c_e=n\}\vert$ be the in-degree of a node $n$,
i.e., the number of edges having $n$ as a child.
Each additional edge that is ancestral to a given node
implies one additional ancestral recombination event,
and thus a factor of  $r^{\deg_i(n)-1}$ in the likelihood,
if the node has any parents.
Put together, these two contribute
\begin{equation}\label{eq:coal}
    \prod_{n \in N} \lambda^{\deg_o(n)-1} r^{\deg_i(n)-1}  
    =
    r^{|E|-|N|+n_r} 
    \lambda^{|E|-|N|+n_s} 
\end{equation}
to the likelihood,
where $n_r$ is the number of nodes with in-degree 0 (roots),
and $n_s$ is the number of nodes with out-degree 0 (usually, samples).
(The simple form follows because each edge contributes exactly one
to some node's out-degree and some node's in-degree.)

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
        Graphical representation of the events associated with an edge $e(p,c,x_c,y)$.
        The lineage carried by $c$ spans $[x,y)$ until a recombination event splits it
        into two segments. The right hand segment coalesces at time $t_p$. The time to
        recombination is thus bounded by $(c. \min(t_p, t_{p'})$.
        \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$.
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times S)$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligbile links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occurr, then the number of links not
affected by recombination is 1 less (see \ref{eq:depth}).
